An Optimized Agent for Capture The Flag Pac-Man

### 2011-03-09 projectWriteup/paper.tex
Fixed some final bugs and finished paper

\begin{document}
\maketitle
\section{Introduction}
A popular and relevant field in artificial intelligence is noisy gameplay
\cite{miik}.  Noisy readouts, opposing agents, and limited calculation
time, mean that noisy games are open to a wide variety of AI techniques.
One particularly relevant example of this field is Capture the Flag (CTF)
Pac-Man, in which a team of independent agents tries to simultaneously
protect food on their side of a Pac-Man board while eating the food on
their opponents' side.  In this paper, we discuss our implementation of a
balanced agent to play CTF Pac-Man.
\\ \\
In our balanced agent, we observed that there are three main branches of
strategy in CTF Pac-Man: attacking, holding the line, and defending.  The
first attempts to eat opposing food, the second attempts to block opposing
agents from entering our side, and the third attempts to capture opposing
agents and prevent them from eating our food once they cross over.  We
built individual modules for each one of these strategies to recommend
moves.  As in many AI agents, the recommendations of these modules and the
global decision of which module to apply use a combination of minimax and
expectimax, where we look several moves ahead, analyze the strength of our
position, and act accordingly.
\\ \\
Our differentiator and the primary focus of our work was on more advanced
ways to model opponents and determine the value of a given game state,
as well as techniques to optimize the weights used to combine those
parameters into relevant state values.  For inference of opponent agent
positions, using prior knowledge and the noisy readings we could observe,
we used more than just sensor data, and incorporated a basic opponent
model.  Our various agents incorporated a wide variety of heuristics,
weighted to recommend moves.  We used a technique called Particle Swarm
Optimization (PSO) to find a string set of weights.  PSO tends to perform
well in functions like ours that are difficult to evaluate accurately and
impossible to analyze from a theoretical perspective \cite{carlisle}.
Furthermore, we observed that the best weightings varied between different
layouts.  Therefore, we trained weights for many different boards.  When
presented with a new board, we clustered it with similar boards (based on
a number of features of the layout) for which
we had trained values and used an average of those values.  The result is
an agent that has learned how to play well on a variety of different
layouts with a variety of different strategies, and combines that
knowledge effectively when presented with a new game.

@@

Hence, we wrote a heuristic function that looked at a board and classified each of the agents, theirs and ours, as being in one of those three states.  To evaluate a position we took a weighted sum of how the three heuristics thought we were doing; thus our heuristic was considering a weighting of how well our defenders were protecting our dots, our attackers were threatening their dots and our hold the line agents were preventing them from crossing onto our side.
\subsection{Offense}
\subsection{Defense}
The defensive heuristics that we found most helpful were putting emphasis on staying close to our enemies, getting them into one exit scenarios and ensuring that for any piece of food we were closer than our opponents.  A key limitation that we had was that losing sight of attacking enemies could be very dangerous because lack of information is very helpful to attackers.  However, because we were able to deduce an exact location any time food went missing, any agent that was doing serious harm was inherently easy to track.
\subsection{Hold the Line}
Playing defense against agents who had not crossed onto our side was a very different problem than playing it against agents who had, mostly because trying to eat them was no longer a strong strategy.  Hence we hit upon the idea of separating that type of play into a ``Hold The Line" heuristic.  For the purposes of holding the line, our agents would ``match up'' against usually one but sometimes more, if we had some agents attacking, opposing agents and mirror their moves, trying to stay closer to all of the squares on the border than the opponents that that agent was ``covering''.  With good inference, as is possible in a 6 agent game, it could become very hard for any opposing agents to cross the border.

\section{Inference}
For inference of opponent agent positions, we update prior probabilities based not only on our sensor readings, but also based on assumptions in our opponent model, which assumes that the opponent is at all times more likely to be moving towards our food and, therefore, places higher probabilities on positions that get them closer. Additionally, our inference module analyzes our food grid and, whenever one of our pieces of food gets eaten, updates the probabilities of the enemy agent that ate it (whichever agent moved before our agent who' is currently analyzing his sensor readings). We are also careful in our inference module about cases where we eat an opponent, to update the probabilities to reflect that they are starting over at their initial position.

@@

Part of the appeal of Particle Swarm Optimization (PSO) is that it has in practice been shown to be fairly robust against noise in evaluating an objective function.  PSO was proposed in 1995 by \cite{originalref} for optimizing functions that are continuous but have no other useful structure.  The appeal of PSO over a monte carlo stochastic optimization scheme such as \cite{montecarlo} is that we have a strong suspicion that our true objective function is locally pretty smooth, which fits nicely with local search based methods like PSO.  Additionally using more particles helps us avoid local minima to which genetic algorithms are susceptible as suggested by \cite{geneticalgorithm}.  However, gradient descent is not an option because we cannot compute a first derivative and empirical gradient descent would be noisy because we cannot accurately compute the function at any one point so our empirical gradients would be very inaccurate.

@@

\FOR {iterations $j$ in $1, ..., k$}
  \FOR{ each particle $i$ in $1, ..., N$}
     \STATE Pick random numbers $r_p, r_g \sim U(0,1)$.
     \STATE Set $v_i = \omega v_i + \omega_p r_p (p_i - x_i) + \omega_g r_g (p -x_i)$.
     \STATE Set $x_i = x_i + v_i$.
     \IF {$x_i >_{agent} p_i$}
       \STATE Set $p_i=x_i$.
       \IF{$p_i >_{agent} p$}
         \STATE Set $p = p_i$.
       \ENDIF
     \ENDIF
  \ENDFOR
\ENDFOR

@@

\subsection{Performance and Results}

\section{$k$-means Clustering for Layouts}
To solve this problem of disparate weights selected by PSO for different layouts, given a layout, we attempted to determine similar layouts for which we had already computed training data.  The idea was that using the single ``closest'' layout would be a mistake since each of our weights were computed by PSO and were thus subject to noise; adding independent results therefore seems likely to result in better performance.

\subsection{Methodology}
We trained 20 different agents on each of the contest layouts.  Additionally we extracted some features from each board - such as amount of food, size and number of agents.  Then we used k-means clustering to choose a handful of clusters that capture the data pretty well.  The choice of k-means clustering was done because it seemed likely that clustering might help as suggested by \cite{textbook} and we consider k-means clustering to be the default for problems without a prior \cite{kmeans}.  Moreover the the centroid of a set of points is the natural choice for how we would mix a set of heuristics, making k-means seem to match the same prior.

@@

\subsection{Performance and Results}

\section{Discussion}

@@

>>>>

\begin{document}
\maketitle
\section{Introduction}
A popular and relevant field in artificial intelligence is noisy gameplay
\cite{miik}.  One fun example of this field is Capture the Flag (CTF)
Pac-Man.  In this paper, we discuss our implementation of an optimized, flexible agent to play CTF Pac-Man.

In our balanced agent, we observed that there are three main branches of
strategy in CTF Pac-Man: attacking, holding the line, and defending.  The
first attempts to eat opposing food, the second attempts to block opposing
agents from entering our side, and the third attempts to capture opposing
agents and prevent them from eating our food once they cross over.  We
built individual modules for each one of these strategies to recommend
moves.  As in many AI agents, the recommendations of these modules and the
global decision of which module to apply use a combination of minimax and
expectimax.

Our differentiator and the primary focus of our work was on more advanced
ways to model opponents and determine the value of a given game state,
as well as techniques to optimize the weights used to combine those
parameters into relevant state values.  We used opponent modeling and clever observations in addition to our noisy readouts to infer our opponents' positions.  Our various agents incorporated a wide variety of heuristics,
weighted to recommend moves.  We used a technique called Particle Swarm
Optimization (PSO) to find a strong set of weights.  PSO tends to perform
well in functions like ours that are both difficult to evaluate accurately and
impossible to analyze from a theoretical perspective \cite{carlisle}.  To integrate the specific results of PSO to new layouts, we clustered a layout at gametime with other layouts for which we had pre-computed PSO values.

\section{Inference}
The first problem we had to solve was that of determining where opposing agents were: the noisy readings from the game state left much to be desired.  For inference of opponent agent positions, we update prior probabilities based not only on our sensor readings, but also based on assumptions in our opponent model, which assumes that the opponent is at all times more likely to be moving towards our food and, therefore, places higher probabilities on positions that get them closer. Additionally, our inference module analyzes our food grid and, whenever one of our pieces of food gets eaten, updates the probabilities of the enemy agent that ate it (whichever agent moved before our agent who is currently analyzing its sensor readings). For obvious reasons, our inference module also handles the two cases in which we know an opponents' location: when we eat them, their probability mass goes back to their start position, and once they are in ``line of sight'' we update their probability distribution accordingly.

@@

Hence, we wrote a heuristic function that looked at a board and classified each of the agents, theirs and ours, as being in one of those three states.  To evaluate a position we took a weighted sum of how the three heuristics thought we were doing; thus our heuristic was considering a weighting of how well our defenders were protecting our dots, our attackers were threatening their dots and our hold the line agents were preventing them from crossing onto our side.
\subsection{Offense}
Playing offense used a number of heuristics designed to reward agents that could capture food maximally on each attack run.  We found that the most important heuristics were the number of food on the board, our distance to the nearest food, and our total distance to food items.  We also included heuristics taking into account how close we were to other agents, both opposing ghosts and teammates.  Finally, we imposed a heuristic penalty for entering cells of the board that were trapped by opposing agents.

\subsection{Defense}
The defensive heuristics that we found most helpful were staying close to our enemies, getting them into single-exit (trapped) scenarios and ensuring that for any piece of food we were closer than our opponents.  A key limitation that we had was that losing sight of attacking enemies could be very dangerous because mutual lack of information is very helpful to attackers.  However, because we were able to deduce an exact location any time food went missing, any agent that was doing serious harm was inherently easy to track.
\subsection{Hold the Line}
Playing defense against agents who had not crossed onto our side was a very different problem than playing it against agents who had, mostly because trying to eat them was no longer a strong strategy.  Hence we hit upon the idea of separating that type of play into a ``Hold The Line" heuristic.  For the purposes of holding the line, our agents would ``match up'' against usually one but sometimes more opposing agents depending on whether we had some agents attacking.  Our agents mirror their moves, trying to stay closer to all of the squares on the border than the opponents that that agent was ``covering''.  With good inference, as is possible in a 6 agent game, it could become very hard for any opposing agents to cross the border.

@@

Part of the appeal of Particle Swarm Optimization (PSO) is that it has in practice been shown to be fairly robust against noise in evaluating an objective function \cite{pars}.  PSO was proposed in 1995 by \cite{originalref} for optimizing functions that are continuous but have no other useful structure.  The appeal of PSO over a Monte Carlo stochastic optimization scheme such as \cite{montecarlo} is that we have a strong suspicion that our true objective function is locally pretty smooth, which fits nicely with local search based methods like PSO.  Additionally using more particles helps us avoid local minima to which genetic algorithms are susceptible as suggested by \cite{geneticalgorithm}.  However, gradient descent is not an option because we cannot compute a first derivative and empirical gradient descent would be noisy because we cannot accurately compute the function at any one point so our empirical gradients would be very inaccurate.

@@

\FOR {iterations $j$ in $1, ..., k$}
  \FOR{ each particle $i$ in $1, ..., N$}
     \STATE Pick random numbers $r_p, r_g \sim U(0,1)$.
     \STATE Set $v_i = \omega v_i + \phi_p r_p (p_i - x_i) + \phi_g r_g (p -x_i)$.
     \STATE Set $x_i = x_i + v_i$.
     \IF {$x_i >_{agent} p_i$}
       \STATE Set $p_i=x_i$.
       \IF{$p_i >_{agent} p$}
         \STATE Set $p = p_i$.
       \ENDIF
     \ENDIF
  \ENDFOR
\ENDFOR

@@

\subsection{Performance and Results}
Overall, we found that on the standard capture map, agents whose weights were selected by PSO tended to perform slightly worse than our hand-optimized agent, tying 61.5\% of the time, losing 30.7\% of the time, and only winning 7.7\% of the time.  In determining the reason for this apparent failure of PSO, we determined that for an average variable optimized by PSO, across all different maps we attempted, the standard deviation was 29.8\% of the possible range, and 39.8\% of the observed range.  In other words, PSO returned very different optimized results for different layouts.  Indeed, PSO won 38.5\% of the time, tied 46.1\% of the time, and lost 15.4\% of the time on the map for which it had been optimized.

\section{$k$-means Clustering for Layouts}
To solve this problem of disparate weights selected by PSO for different layouts, given a layout, we attempted to determine similar layouts for which we had already computed training data.  The idea was that using the single ``closest'' layout would be a mistake since each of our weights were computed by PSO and were thus subject to noise; averaging independent results therefore seems likely to result in better performance.

\subsection{Methodology}
We trained 20 different agents on each of the contest layouts.  Additionally we extracted some features from each board - such as amount of food, size and number of agents.  Then we used $k$-means clustering to choose a cluster similar to our observed layout.  We chose $k$-means as clustering can stabilize disparate data \cite{textbook} and we consider $k$-means the default for problems without a prior \cite{kmeans}.  Moreover the the centroid of a set of points is the natural choice for how we would mix a set of heuristics, and $k$-means matches the same metric.

@@

\subsection{Performance and Results}
Overall, clustering significantly improved the efficacy of our PSO-optimized weights.  In comparison to our hand-optimized agent, among games that one of the two agents won, our averaged PSO-weighted agent won approximately 64\% of the games.  The data suggests that clustering was able to mitigate the damage of optimizing to the wrong board, and allow us to better approximate an optimal weighting.

\section{Discussion}
Our approach to Pac-Man consisted of a mix of human-determined approaches and machine learning.  We began with a collection of agents that were primarily human-optimzied, using pre-determined weights on a subset of the heuristics described above for evaluation on a Minimax tree.  These agents was only marginally more successful than the reflex agents, and had several key shortcomings; namely, that they were unable to coordinate -- to choose offense or defense -- and lacked optimal parameters.  We solved the first problem initially by assigning weights to a series of board meta-analysis conditions to switch between agent recommendations.  This substantially improved the robustness of our player, but our hand-optimized weights were still quite weak.  Our inital approach to use binary Particle Swarm Optimization to optimize these parameters was ineffective in practice, because PSO tended to give heavily board-specific weights with high variance between boards.  Applying $k$-means clustering was effective in concert with PSO to ensure that the weights suggested by our particle swarm were effectively applied to a given layout.

Overall, our agent's strongest asset is its flexibility -- a result of a wide variety of heuristics generated from human analysis of the game, PSO to develop correct weightings across different maps, and clustering to determine the proper gametime weighting.  Continued improvement could be achieved by using standard PSO, rather than binary PSO, assuming we had some objective function to minimize.  Such a function would likely be an objective measure of percentage success against a static cross-section of agents, such as repeated participation in a nightly tournament.  Additionally, running PSO and $k$-means on a larger set of layouts would likely give our agent more flexibility and encourage better performance on nonstandard layouts.

@@

\bibitem{pars} Parsopoulos, K. E., Vrahatis, M. N. (2001) Particle Swarm Optimizer in Noisy and Continuously Changing Environments. {\it Proceeding of the IASTED International Conference on Artificial Intelligence and Soft Computing} (ASC 2001), Cancun, Mexico. pp. 289-294.
